Retrofitting Effect Handlers onto OCaml

### 2020-11-12 retro-concurrency.tex
sec1

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our
goal is to retrofit effect handlers onto the OCaml programming language, which
has been in continuous use for the past 25 years in large codebases including
verification tools~\cite{everest,Coq}, mission critical software
systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with fast foreign-function interface
(FFI) and excellent compatibility with program analysis tools such as debuggers
and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to obtain a
backtrace. Our primary motivation for adding effect handlers to OCaml is to
support (1) highly concurrent applications such as web servers using a
natural \emph{direct-style} as opposed to using callbacks and (2) efficient
nested parallel programs in the vein of Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP},
X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and variants of Parallel
ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then
reifying continuations in the source language is relatively straight-forward.
But OCaml does not use CPS IR, and changing the OCaml compiler to utilise CPS IR
would be an enormous undertaking. Utilizing a CPS IR would also affect the
performance profile of OCaml applications due to increased memory allocations as
continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
with CPS, stack unwinding is meaningless and hence, we would lose compatibility
with tools that inspect the program stack. Hence, we choose not to use CPS
translation and represent continuations as call stacks.

>>>>

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts, our
goal is to retrofit effect handlers onto the OCaml programming language, which
has been in continuous use for the past 25 years in large codebases including
verification tools~\cite{everest,Coq}, mission critical software
systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with a fast foreign-function interface
(FFI). It has excellent compatibility with program analysis tools such as debuggers
and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to obtain a
backtrace. Our primary motivation for adding effect handlers to OCaml is to
support: {\em (i)} highly concurrent applications such as web servers using a
natural \emph{direct-style} as opposed to using callbacks; and {\em (ii)} efficient
nested parallel programs in the vein of Cilk~\cite{Cilk}, OpenMP~\cite{OpenMP},
X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and variants of Parallel
ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then
reifying continuations in the source language is relatively straight-forward.
But OCaml does not use a CPS IR, and changing the OCaml compiler to utilise a CPS IR
would be an enormous undertaking that would affect the
performance profile of existing OCaml applications due to increased memory allocations as
continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
with CPS, stack unwinding is meaningless and hence, we would lose compatibility
with tools that inspect the program stack. Hence, we choose not to use CPS
translation and represent continuations as call stacks.
